Rapid and Accurate Global PTM Discovery (G-PTM-D) Using Post-Acquisition Spectral Calibration and Multi-Notch Searches

### 2017-10-20 paper.tex
nice

The described issues are apparent in mass difference histograms in Figure~\ref{fig:fig4mouse-1012}, generated by compiling the results of a 2.5 Da tolerance narrow-window search.
These plots clearly show that numerous peptides have missed monoisotopic (Mm) peaks, deamidation, or an amino acid substitution.

@@

\begin{table}[]
\centering
\caption{Notch Search vs Narrow-Window Search comparison of PSMs for Mouse}
\label{tbl:singleVsMultiNotch}
\begin{tabular}{llll}
                    & 5ppm Search & 3.5 Da Search & Notch Search \\
PSMs Within 1\% FDR & 157188      & 178003        & 177213       \\
misassigned PSMs    & 0           & 1521          & 0            \\
\end{tabular}
\end{table}

@@

There is a major problem with using the results from unrestricted precursor mass tolerance searches to infer peptide modifications regardless of whether they are of the Chick or Kong variety.
The freedom in precursor mass difference results in a multitude of false positive \textit{mass difference identifications}.
For example, small missed cleavages are often reported as mass differences.
The peptide PEPKK is often identified as PEPK with a mass error of 128 Da (the converse is also true, PEPK may be identified with a mass error of -128 Da).
This error is relatively easy to spot but when the missed cleavages contain multiple amino acids, the interpretation and discovery of the error is much more difficult.
Another example of this relates to peptide modifications.
The phosphopeptide PEPpSK could be reported with a mass error of -79.96 Da when the better identification would have been PEPSK.
Constraining the search using the multi-notch approach described below, directly prohibits erroneous assignment of the best match for a peak, and instead produces an alternative peak assignment with a much lower false positive rate (see below).

@@

[Cite Bo Zhang if paper exists].
We showed above in some amazing figure an example where several peptides were co-eluting from the column into that mass-spec within the tolerance used for isolation for fragmentation.
In that particular case, the precursor with m/z dsfadf was chosen for isolation and fragmentation.
In a conventional search, only that precursor mass would be used to select a range of theoretical for comparison.
However, all peptides in that range are co-fragmented and a better solution would be to use all co-isolated precursor masses.
In MetaMorpheus, for each MS2 scan, the corresponding MS1 scan is deconvoluted and the complete list of precursors is gathered for the range surrounding the isolated precursor ($\pm 3$ Th).
Each deconvoluted mass is used to select a range of theoreticals to use for identification.
The result of this is that often, more than one peptide can be identified from a single MS2 scan.
Below, in Table 3434, we should the number of PSMs for a raw file where we use only a single precursor mass and we compare that to the case where we allow for multiple different precursors.
You can see a significant increase in the number of identified PSMs.
Importantly, you can see in Supplemental Figure erf3e, that the precursor mass errors go down significantly and the standard deviation is much reduced.
This is extremely beneficial in both open mass and multi-notch searches because it reduces the number the more difficult to interpret mass shifts and substantially improves G-PTM-D (see below).

>>>>

The described issues are apparent in mass difference histograms in Figure~\ref{fig:fig4jurkat-1012}, generated by compiling the results of a 2.5 Da tolerance narrow-window search.
These plots clearly show that numerous peptides have missed monoisotopic (Mm) peaks, deamidation, or an amino acid substitution.

\begin{figure}[H]
\includegraphics{fig4jurkat-1012.png}
\caption{Histograms PSM mass differences found in various dalton ranges}
\label{fig:fig4jurkat-1012}
\end{figure}

@@

\begin{table}[]
\centering
\caption{Notch Search vs Narrow-Window Search comparison of PSMs for Mouse}
\label{tbl:singleVsMultiNotch}
\begin{tabular}{llll}
                    & 5ppm Search & 2.5 Da Search & Notch Search \\
PSMs Within 1\% FDR & 275359      & 262543        & 279895       \\
misassigned PSMs    & ?           & ?          & ?            \\
\end{tabular}
\end{table}

A specific example of a misassigned PSM is the example where the correct identification has oxidation and ammonia loss, but was identified as an unmodified peptide with score 10.
With G-PTM-D added modifications, it is clear that the correct identification has a mass difference of 0 and a score 20.

@@

There is a major problem with using the results from unrestricted precursor mass tolerance searches to infer peptide modifications regardless of whether they are of the Chick or Kong variety.
The freedom in precursor mass difference results in a multitude of false positive \textit{mass difference identifications}.
For example, small missed cleavages are often reported as mass differences.
The peptide PEPKK is often identified as PEPK with a mass error of 128 Da (the converse is also true, PEPK may be identified with a mass error of -128 Da).
This error is relatively easy to spot but when the missed cleavages contain multiple amino acids, the interpretation and discovery of the error is much more difficult.
Another example of this relates to peptide modifications.
The phosphopeptide PEPpSK could be reported with a mass error of -79.96 Da when the better identification would have been PEPSK.

@@

Consider the scan JurkatFrac12 scan 12336. Isolation m/z  = 726.65.
Isolation width is 2.5 Th.
It is shown in Figure~\ref{fig:fig5-coIsolationSpectrum}.

\begin{figure}[H]
\includegraphics{fig5-coIsolationSpectrum.png}
\caption{m/z peaks in scan 12333}
\label{fig:fig5-coIsolationSpectrum}
\end{figure}

We showed above in some amazing figure an example where several peptides were co-eluting from the column into that mass-spec within the tolerance used for isolation for fragmentation.
In that particular case, the precursor with m/z dsfadf was chosen for isolation and fragmentation.
In a conventional search, only that precursor mass would be used to select a range of theoretical for comparison.
However, all peptides in that range are co-fragmented and a better solution would be to use all co-isolated precursor masses.
In MetaMorpheus, for each MS2 scan, the corresponding MS1 scan is deconvoluted and the complete list of precursors is gathered for the range surrounding the isolated precursor ($\pm 3$ Th).
Each deconvoluted mass is used to select a range of theoreticals to use for identification.
The result of this is that often, more than one peptide can be identified from a single MS2 scan.
Below, in Table 3434, we should the number of PSMs for a raw file where we use only a single precursor mass and we compare that to the case where we allow for multiple different precursors.
You can see a significant increase in the number of identified PSMs.
Importantly, you can see in Supplemental Figure erf3e, that the precursor mass errors go down significantly and the standard deviation is much reduced.
This is extremely beneficial in both open mass and multi-notch searches because it reduces the number the more difficult to interpret mass shifts and substantially improves G-PTM-D (see below).
